---
title: "EIX: Explain Interactions in XGBoost"
author: "Ewelina Karbowiak"
date: "2018-12-07"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{EIX Explain Interactions in XGBoost}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, echo=FALSE}
knitr::opts_chunk$set(fig.height = 6, fig.width = 6, fig.align = "center")

```

Package `EIX` is the set of tools to explore the structure of XGBoost and lightGBM models. It includes functions finding strong interactions and also checking importance of single variables and interactions by usage different measures. `EIX` consists several functions to visualize results.

Almost all `EIX` functions require only two parameters: a XGBoost or LightGBM model and data table used as training dataset. The exceptions are the `waterfall` function and its plot. The first one requires parameters: a XGBoost model and observation, which prediction has to be explained). These two functions support only XGBoost models. All plots are created with package `ggplot2`. Most of them use plot theme `theme_mi2` from `DALEX`.

## Data Info

This vignette shows usage of `EIX` package. It lets to explain XGBoost prediction model concerning departures of employees from company using HR_data. Dataset was taken from kaggle and consists 14999 observations and 10 variables. The dataset is also available in package `EIX` and there it is described more precisely.

```{r}
#devtools :: install_github("ModelOriented/EIX")
library("EIX")
set.seed(4)
knitr::kable(head(HR_data))
```

To create correct XGBoost model, remember to change categorical features to factors and next change the data frame to sparse matrix. The categorical features are one-hot encoded.

```{r, warning=FALSE, message=FALSE}
library("Matrix")
sparse_matrix <- sparse.model.matrix(left ~ . - 1,  data = HR_data)
head(sparse_matrix)
```


## Xgboost model creation

Package `EIX` uses table, which was generated by `xgboost::xgb.model.dt.tree` with information about trees, their nodes and leaves.

```{r}
library("xgboost")
param <- list(objective = "binary:logistic", max_depth = 2)
xgb_model <- xgboost(sparse_matrix, params = param, label = HR_data[, left] == 1, nrounds = 50, verbose = FALSE)
knitr::kable(head(xgboost::xgb.model.dt.tree(colnames(sparse_matrix),xgb_model)))
```

Function `xgboost::xgb.importance` shows importance of single variables. `EIX` adds new measures of variables’ importance and shows also importance of interactions.

```{r}
knitr::kable(head(xgboost::xgb.importance(colnames(sparse_matrix),xgb_model)))
```

## Model visualization

The `lollipop` plot is used to visualize the model in such way that the most important variables and interactions are visible. 

On the x-axis, there are tree numbers and on the y-axis there is **Gain** measure for each node. One segment is one tree in the model and each point is one node. On the plot there are all nodes, which are not leaves. Shape of points signifies depth of node. All roots on the plot are connected by a red line. If in the same segment there is a variable with a higher depth above the variable with a lower depth, it means that interaction occurs.

There is opportunity to set a different way of labeling. On the plot we can see the most important variables in roots (horizontal labels), and interactions (vertical labels), this is option `labels = "topAll"` which is default. Moreover, there are two additional options: `labels = "roots"` - for variables in roots only, `labels = "interactions"` for interactions only. The numbers of labels visible on the plot you can change by parametr `threshold` (range from 0 to 1, default 0.1).
The plot is on a logarithmic scale because the initial trees usually are the most important. You can change the scale of the plot by setting the parameter `log_scale = FALSE`.

```{r}
lolli<-lollipop(xgb_model,sparse_matrix)
plot(lolli)
#plot(lolli, threshold=0.05)
#plot(lolli, labels="roots")
#plot(lolli, labels="interactions")
#plot(lolli, labels="roots", threshold=0.05)
#plot(lolli, labels="interactions",threshold=0.05)
#plot(lolli, log_scale = FALSE)
```


## Interactions

We can consider interactions in two ways. In first approach we can explore all pairs of variable, which occur in the model one above the other. This approach is not the best one, because we cannot distinguish if pair of variables are real interaction or not. In this approach high gain of pair can be a result of high gain of down variable (child). To explore pairs of variables you can generate table with them using function `interactions` with parametr `option = "pairs"`. This table includes **Gain** measure and  number of occurrences of pairs. You can also use the function `plot` to visualize **Gain** measure.

```{r}
pairs<-interactions(xgb_model, sparse_matrix, option = "pairs")
head(pairs)
plot(pairs)
```

The `interactions` plot is a matrix plot with a child from the pair on the x-axis and the parent on the y-axis. The color of the square at the intersection of two variables means value of **sumGain** measure. The darker square, the higher **sumGain** of variable pairs. The range of **sumGain** measure is divided into four equal parts: `very low, low, medium, high`.

In second approach, to find strong interactions, we can consider only these pairs of variables, where variable on the bottom (child) has higher gain than variable on the top (parent). We can also create ranking of interactions using function `importance` with parameter `option = "interactions"`. More details in the next section.

```{r, warning=FALSE, message=FALSE}
interactions<-interactions(xgb_model, sparse_matrix, option = "interactions")
head(interactions)
plot(interactions)
```


## Variables' and interactions’ importance

For exploring variables’ and interactions’ importance there are three functions in `EIX` package: `importance`, its `plot` with parameter `radar = TRUE` or `radar = FALSE`.
With `EIX` package we can compare importance of single variables and interactions. The functions `importance` can return three kinds of outputs, depending on the `opt` parameter:

 * `option = "variables"` - it consists only single variables

 * `option = "interactions"`- only interactions

 * `option = "both"`-  output shows importance both single variables and interactions.
 
 NOTE: `option = "both"` is not direct connection `option = "variables"` and `option = "interactions"`, because values of variable importance measure, which were in the interactions, are not included in importance of single variable.

In `EIX` the following measures are available:

* **sumGain** - sum of Gain value in all nodes, in which given variable occurs
* **sumCover** - sum of Cover value in all nodes, in which given variable occurs; for LightGBM models: number of observation, which pass through the node
* **mean5Gain** - mean gain from 5 occurrences of given variable with the highest gain
* **meanGain** - mean Gain value in all nodes, in which given variable occurs
* **meanCover** - mean Cover value in all nodes, in which given variable occurs; for LightGBM models: mean number of observation, which pass through the node
* **freqency** - number of occurrences in the nodes for given variable

`EIX` package gives additionally measures of variables importance for single variable:

* **counterRoot** - number of occurrences in the root
* **weightedRoot**  - mean number of occurrences in the root, which is weighted by gain
* **meanDepth ** - mean depth weighted by gain

The function `importance` returns a table with all available importance measures for given option. 
The table is sorted by descending value of **sumGain**.

The function `plot` with parameter  `radar = FALSE` and a result from the `importance` function as an argument  shows two measures of importance, which can be chosen by `xmeasure` and `ymeasure` parameters. By parameter `top` we can decide how many positions will be included in the plot.

```{r}
importance<-importance(xgb_model, sparse_matrix, option = "both")
head(importance)
plot(importance, radar=FALSE)
#plot(importance,  xmeasure = "mean5Gain", ymeasure = "sumGain",  top = 15, radar=FALSE)
```

The function `plot` with parameter `radar = TRUE` enables to compare different measures of variables and interactions importance on the radar plot from `ggiraphExtra` package.
Bellow I attach the example of radar plot. On the outside of the circle there are names of variables or interactions. Colored lines represent various measures  of importance. The positions on the plot are sorted decreasing. The variable with the highest **sumGain** value is on the right of 12 o'clock. Next the **sumGain** value decreases in a clockwise direction.  On the plot it is possible to change place, where the features names start by parameter `text_start_point` (range from 0 to 1, default 0.5), and size of this text by parametrer `text_size`.

```{r}
plot(importance)
#plot(importance, text_start_point = 0.3)
#plot(importance, text_size = 4)
#plot(importance, top=15)
```

## Explanation of single prediction including interactions

For single prediction explaining package `EIX` uses two packages: `xgboostExplainer` i `breakDown`. The package `xgboostExplainer` is a tool to interpreting prediction of xgboost model. The package `EIX` uses its code and modifies it to include interactions. The methodology of plot creation comes from package `breakDown`.

The function `waterfall` returns table with variables’ impact on the prediction of the model.  Depending on the parameter `option`, the table includes interactions (`option = "interactions"`- default) or does not (`option = "variables"`). The function `plot` with `waterfall` object as an argument visualizes this table. On the y-axis there are: intercept (it is the probability that random variable from training dataset will be 1), variables (which have an impact on prediction) and final prognosis of the model. On the x-axis there is log-odds of impact each variables.

```{r}
data <- HR_data[9,]
new_observation <- sparse_matrix[9,]
wf<-waterfall(xgb_model, new_observation, data, option = "interactions")
wf
plot(wf)
#wf<-waterfall(xgb_model, new_observation, data, option = "interactions", baseline = "intercept")
#wf
#plot(wf)
```
