---
title: "EIX: Titanic data "
author: "Ewelina Karbowiak"
date: "2018-29-03"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{EIX for Titanic data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, echo=FALSE}
knitr::opts_chunk$set(fig.height = 6, fig.width = 6, fig.align = "center")

```

## Data Info
This vignette shows usage of `EIX` package for titanic data. This dataset was copied from `stablelearner` package. With `EIX` package we explain XGBoost classification model concerning the survival problem. More details about `EIX` package [here](https://modeloriented.github.io/EIX/articles/EIX.html).


```{r}
#devtools :: install_github("ModelOriented/EIX")
library("EIX")
library(data.table)
set.seed(4)
titanic_data<-data.table(na.omit(titanic_data))
knitr::kable(head(titanic_data))

library("Matrix")
sparse_matrix <- sparse.model.matrix(survived ~ . - 1,  data = titanic_data)
```


## Xgboost model creation


```{r, warning=FALSE, message=FALSE}
library("xgboost")
param <- list(objective = "binary:logistic", max_depth = 2)
xgb_model <- xgboost(sparse_matrix, params = param, label = titanic_data[, "survived"] == "yes", nrounds = 50, verbose = FALSE)
```


## Model visualization

First let's plot the model.

```{r}
lolli<-lollipop(xgb_model,sparse_matrix)
plot(lolli, threshold=0.02)
```


## Interactions

Next we explore interactions using `interactions()` functions and its plot.

```{r}
interactions<-interactions(xgb_model, sparse_matrix, option = "interactions")
head(interactions, 15)
plot(interactions)
```

## Variables' and interactions’ importance

```{r}
importance<-importance(xgb_model, sparse_matrix, option = "both")
head(importance, 15)
plot(importance, radar=FALSE)
```

```{r}
plot(importance)
```

## Explanation of the single prediction including interactions

Let's see an explanation of the prediction for an 18-year-old from England who has traveled 3rd class.

```{r}
data <- titanic_data[27,]
new_observation <- sparse_matrix[27,]
wf<-waterfall(xgb_model, new_observation, data, option = "interactions")
wf
plot(wf)
```
